---
title: "Lab 1 - Data Analysis and Modeling"
author: "Yue Zhang & SiHan Chen"
date: "3/25/2025"
output:
  html_document:
    df_print: paged
  pdf_document:
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This lab introduces the basics of R and RStudio.
We will work with two built-in datasets, pressure and women, to practice data exploration, visualization, and analysis. We will complete 7 exercises and generate reproducible results in R Markdown format.

# Exercise 1
Task: Complete the R language basics 1_3 from Cookbook for R _ Basics
Status: [OK] Done


# Exercise 2 - Exploring the pressure dataset
```{r load-pressure}
data(pressure)
head(pressure)
```
## Select specific data:

Fourth row, second column:
```{r select-point}
pressure[4, 2]
```
Select the entire second column:

```{r select-column}
pressure[, 2]
```


# Exercise 3 - Temperature vs. Pressure Plot
```{r pressure-plot}
plot(
  x = pressure$temperature,
  y = pressure$pressure,
  type = "b",
  col = "darkred",
  pch = 19,
  lwd = 2,
  xlab = "Temperature (°C)",
  ylab = "Pressure (mmHg)",
  main = "Temperature vs Pressure"
)
grid()
```
From the plot, we observe that as temperature increases, pressure also increases rapidly, showing a non-linear relationship.


# Exercise 4 - Calculating Volume using PV = nRT
We use the ideal gas law:
$$PV = nRT$$
Where:
	•	( P ): pressure in Pa
	•	( T ): temperature in Kelvin
	•	( V ): volume
	•	( R = 8.31 ), ( n = 1 )
```{r calc-volume}
R <- 8.31
n <- 1
T <- pressure$temperature + 273.15
P <- pressure$pressure * 133.322  # mmHg to Pascal

V <- (n * R * T) / P
```

## Plot: Temperature vs Volume
```{r plot-volume}
plot(
  x = pressure$temperature, y = V,
  type = "o", col = "steelblue", pch = 17, lwd = 2,
  xlab = "Temperature (°C)", ylab = "Volume (L)",
  main = "Estimated Volume vs Temperature"
)
grid()
```
This plot demonstrates the volume increases with temperature, as expected from gas law behavior.


# Exercise 5 - Exploring the women dataset
```{r load-women}
data(women)
head(women)
dim(women)
names(women)
```
The dataset contains 15 rows and 2 columns: height (inches), weight (lbs)


# Exercise 6 - Plot Height² vs Weight
```{r high2-plot}
height_sq <- women$height^2

plot(
  x = women$weight, y = height_sq,
  type = "p", pch = 19, col = "forestgreen",
  xlab = "Weight (lbs)", ylab = "Height² (in²)",
  main = "Height Squared vs Weight"
)
abline(lm(height_sq ~ women$weight), col = "darkgray", lwd = 2)
grid()
```
We see a clear upward trend: as weight increases, squared height increases linearly.


# Exercise 7 - BMI Calculation
Formula:
$$ \text{BMI} = \frac{\text{weight (kg)}}{[\text{height (m)}]^2} $$
```{r bmi-calc}
# Convert units
weight_kg <- women$weight * 0.453592
height_m <- women$height * 0.0254

BMI <- weight_kg / (height_m^2)
BMI
```

## Plot BMI by individual
```{r plot-bmi}
barplot(
  BMI,
  col = "tomato",
  names.arg = 1:15,
  main = "BMI of Each Woman",
  xlab = "Woman (Index)", ylab = "BMI"
)
abline(h = 25, col = "blue", lty = 2)
text(x = which.max(BMI), y = max(BMI) + 1, labels = "Highest", col = "darkred")
```

```{r max-bmi}
max_bmi_index <- which.max(BMI)
max_bmi_value <- max(BMI)
list("Index" = max_bmi_index, "Max BMI" = max_bmi_value)
```
Woman number r max_bmi_index has the highest BMI: r round(max_bmi_value, 2).

Reminder: Healthy BMI range is 18.5/text{-}24.9. This person may exceed that range.